# Lesson 3: Genotype QC and Population Structure

:::cdi-message
- **ID:** GWAS-L03
- **Type:** Conceptual + Implementation
- **Audience:** Public
- **Theme:** Variant filtering, sample QC, and structure diagnostics
:::

## Why Genotype QC Matters

Association testing assumes:

- Genotypes are measured reliably
- Variants are polymorphic
- Missingness is not systematic
- Population structure is properly characterized

If these assumptions fail, false positives can arise before modeling even begins.

Genotype QC is not a preprocessing formality.
It is statistical risk control.

---

## Load Genotype Matrix

For demonstration, we use a simulated genotype matrix where:

- Rows = samples
- Columns = SNPs
- Values = 0, 1, 2 (allele dosage)

```{r load-genotype-matrix}
#| message: false
#| warning: false

geno <- read.csv("data/demo-genotypes.csv", row.names = 1)
dim(geno)
geno[1:5, 1:6]
```

---

## Missingness Per SNP

Variants with excessive missing data can bias association tests.

```{r snp-missingness}
#| message: false
#| warning: false

snp_missing <- colMeans(is.na(geno))
summary(snp_missing)
```

A typical threshold in practice might be 2–5% missingness.
For this demo, we filter SNPs with >5% missingness.

```{r filter-snp-missingness}
#| message: false
#| warning: false

geno_filtered <- geno[, snp_missing <= 0.05]
dim(geno_filtered)
```

---

## Minor Allele Frequency (MAF)

Very rare variants can:

- Inflate variance estimates
- Produce unstable effect sizes
- Generate extreme p-values due to sparse counts

Compute allele frequency and MAF.

```{r compute-maf}
#| message: false
#| warning: false

allele_freq <- colMeans(geno_filtered, na.rm = TRUE) / 2
maf <- pmin(allele_freq, 1 - allele_freq)

summary(maf)
```

Filter SNPs with MAF < 0.05.

```{r filter-maf}
#| message: false
#| warning: false

geno_filtered <- geno_filtered[, maf >= 0.05]
dim(geno_filtered)
```

Interpretation:

Removing very rare variants increases model stability in small datasets.
In large biobank-scale GWAS, rare variant strategies differ and require specialized modeling.

---

## Sample-Level Missingness

Samples with excessive missing genotypes can distort association signals.

```{r sample-missingness}
#| message: false
#| warning: false

sample_missing <- rowMeans(is.na(geno_filtered))
summary(sample_missing)
```

Filter samples with >5% missingness.

```{r filter-sample-missingness}
#| message: false
#| warning: false

keep_samples <- sample_missing <= 0.05
geno_filtered <- geno_filtered[keep_samples, ]
dim(geno_filtered)
```

---

## Principal Component Analysis (PCA)

Population structure is estimated from genotype variation.

Before PCA, we use simple mean imputation to avoid dropping large parts of the genotype matrix due to missingness.
This is a pragmatic step for structure estimation, not a claim about true genotypes.

```{r impute-missing-genotypes}
#| message: false
#| warning: false

geno_mat <- as.matrix(geno_filtered)

for (j in seq_len(ncol(geno_mat))){
  idx <- is.na(geno_mat[, j])
  if (any(idx)){
    mu <- mean(geno_mat[, j], na.rm = TRUE)
    geno_mat[idx, j] <- mu
  }
}

stopifnot(!anyNA(geno_mat))
```

For teaching stability and speed, we run PCA on a subset of SNPs.

```{r thin-snps-for-pca}
#| message: false
#| warning: false

set.seed(1)
m_use <- min(1000, ncol(geno_mat))
keep_snps <- sample(seq_len(ncol(geno_mat)), size = m_use)
geno_mat_pca <- geno_mat[, keep_snps, drop = FALSE]
dim(geno_mat_pca)
```

Scaling can fail if any SNP has zero variance (all values identical).
We drop such SNPs before PCA.

```{r drop-zero-variance-snps}
#| message: false
#| warning: false

sds <- apply(geno_mat_pca, 2, stats::sd)

keep_var <- is.finite(sds) & sds > 0
geno_mat_pca <- geno_mat_pca[, keep_var, drop = FALSE]

dim(geno_mat_pca)
```

```{r run-pca}
#| message: false
#| warning: false

geno_scaled <- scale(geno_mat_pca)

stopifnot(!anyNA(geno_scaled))
stopifnot(all(is.finite(geno_scaled)))

pca <- prcomp(geno_scaled, center = FALSE, scale. = FALSE)

summary(pca)$importance[2, 1:5]
```

---

## Visualize PC1 vs PC2

```{r plot-pca}
#| message: false
#| warning: false

source("scripts/R/cdi-plot-theme.R")
library(ggplot2)

pca_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2]
)

pal <- cdi_palette()

ggplot(pca_df, ggplot2::aes(x = PC1, y = PC2)) +
  ggplot2::geom_point(
    color = pal$teal,
    alpha = 0.85,
    size = 2.4
  ) +
  ggplot2::labs(
    title = "Population structure (PCA)",
    subtitle = "Principal components capture ancestry-related variation",
    x = "PC1",
    y = "PC2"
  ) +
  cdi_theme()
```

### Interpretation

Principal components summarize correlated genotype variation.

Clusters or gradients suggest:

- Ancestry differences
- Batch effects
- Subpopulation structure

Ignoring population structure can produce systematic association inflation.

---

## What QC Achieved

After QC, we have:

- Removed poorly genotyped variants
- Excluded unstable rare variants
- Filtered low-quality samples
- Estimated ancestry structure

Association testing now operates on a more stable genotype matrix.
