# Lesson 4: Association Testing Concepts

:::cdi-message
- **ID:** GWAS-L04
- **Type:** Statistical Foundations
- **Audience:** Public
- **Theme:** Per-SNP testing, adjustment, and interpretation boundaries
:::

## What Association Testing Actually Does

In GWAS, we test whether a genetic variant is associated with a trait.

For a quantitative trait, the simplest model is a linear regression:

Trait ~ SNP + Covariates

The SNP is usually coded as 0, 1, 2, representing allele dosage.

The output provides:

- an effect size (beta)
- a standard error
- a p-value

A small p-value indicates evidence of association, not proof of causality.

---

## Load Analysis Inputs

In this lesson, we combine:

- phenotype and covariates (from Lesson 2)
- a genotype matrix (from the demo data)

```{r setup-and-load-inputs}
#| message: false
#| warning: false

source("scripts/R/cdi-plot-theme.R")
library(ggplot2)

pheno <- read.csv("data/demo-phenotype.csv", stringsAsFactors = FALSE)
covar <- read.csv("data/demo-covariates.csv", stringsAsFactors = FALSE)

df <- merge(pheno, covar, by = "sample_id")

geno <- read.csv("data/demo-genotypes.csv", row.names = 1)

dim(df)
dim(geno)
```

---

## Inspect the Analysis Table

Before fitting any models, confirm the analysis table contains what you think it contains.

```{r inspect-analysis-table}
#| message: false
#| warning: false

str(df)
head(df)
summary(df$trait)
```

You should see the core fields used throughout this guide:

- sample_id
- trait
- age
- sex
- PC1
- PC2

If these are missing or mis-typed, association testing may still run, but interpretation becomes unreliable.

---

## Align Samples

Genotype data uses row names as sample IDs.
We must ensure phenotype rows match genotype rows.

```{r align-samples}
#| message: false
#| warning: false

common_ids <- intersect(df$sample_id, rownames(geno))

df <- df[df$sample_id %in% common_ids, ]
geno <- geno[common_ids, , drop = FALSE]

# Match order exactly
df <- df[match(common_ids, df$sample_id), ]

stopifnot(all(df$sample_id == rownames(geno)))

dim(df)
dim(geno)
```

---

## Minimal Adjustment Model

In Lesson 2, we motivated adjustment for age, sex, and principal components.

In this demo, we use:

Trait ~ age + sex + PC1 + PC2

Association testing adds one SNP at a time to this baseline model.

---

## Example: One SNP Test

To make the process concrete, test a single SNP and interpret the output.

```{r one-snp-test}
#| message: false
#| warning: false

snp_name <- colnames(geno)[1]
x <- geno[, snp_name]

# Mean impute for this SNP (simple demo approach)
if (any(is.na(x))){
  x[is.na(x)] <- mean(x, na.rm = TRUE)
}

fit <- lm(trait ~ x + age + sex + PC1 + PC2, data = df)
summary(fit)$coefficients["x", ]
```

### Interpretation

- The coefficient for x is the estimated trait change per allele.
- The p-value tests whether that coefficient differs from zero.

This is an association test.
It is not yet a biological claim.

---

## Scaling Up: Per-SNP Regression

GWAS repeats the same model across many SNPs.

We will create a small function that returns:

- beta
- standard error
- p-value

```{r snp-test-function}
#| message: false
#| warning: false

test_snp_lm <- function(x, df){
  if (any(is.na(x))){
    x[is.na(x)] <- mean(x, na.rm = TRUE)
  }

  fit <- lm(trait ~ x + age + sex + PC1 + PC2, data = df)
  co <- summary(fit)$coefficients

  c(
    beta = unname(co["x", "Estimate"]),
    se = unname(co["x", "Std. Error"]),
    p = unname(co["x", "Pr(>|t|)"])
  )
}
```

Now apply it across all SNPs.

```{r run-per-snp-gwas}
#| message: false
#| warning: false

res <- t(apply(geno, 2, test_snp_lm, df = df))
res <- as.data.frame(res)
res$snp <- rownames(res)

res <- res[, c("snp", "beta", "se", "p")]
head(res)
```

---

## Multiple Testing

A GWAS typically tests hundreds of thousands to millions of SNPs.

With so many tests:

- some p-values will be small by chance
- naive thresholds (like 0.05) are not meaningful

This is why GWAS uses stringent thresholds (often around 5e-8 in real studies).

In this demo, we focus on concepts rather than strict genome-wide significance.

---

## Effect Size vs Significance

A small p-value can occur with:

- a large effect size
- a small standard error
- a large sample size

A large effect size can be non-significant if sample size is small or variance is large.

Effect size answers:

How strong is the association?

P-value answers:

How surprising is the association under the null?

Both must be interpreted together.

---

## Quick Diagnostic: Top Hits

```{r top-hits}
#| message: false
#| warning: false

res <- res[order(res$p), ]
head(res, 10)
```

At this stage, a top hit is not yet a biological claim.
It is a candidate signal that needs validation and calibration.

---

## What This Lesson Adds to the Reasoning Chain

Association testing provides statistical evidence.
It does not provide mechanism.

Before interpreting a hit, we still need to ask:

- Is the signal robust to covariate adjustment?
- Is it driven by structure or outliers?
- Is the effect size plausible?
- Is it likely replicated, or just a chance finding?

Those checks start in Lesson 5 with visualization and diagnostics.
