# Lesson 5: Visualizing and Validating Association Signals

:::cdi-message
- **ID:** GWAS-L05
- **Type:** Diagnostics + Interpretation
- **Audience:** Public
- **Theme:** Manhattan plots, QQ plots, and inflation assessment
:::

## Why Visualization Matters

Association testing produces a table of p-values and effect sizes.

But raw tables do not reveal:

- global inflation
- systematic bias
- signal concentration
- whether top hits stand out from background noise

Visualization turns a list of p-values into a structure we can interpret.

---

## Recompute GWAS Results

We regenerate results to ensure this lesson is self-contained.

```{r setup-and-run-gwas}
#| message: false
#| warning: false

source("scripts/R/cdi-plot-theme.R")
library(ggplot2)

pheno <- read.csv("data/demo-phenotype.csv", stringsAsFactors = FALSE)
covar <- read.csv("data/demo-covariates.csv", stringsAsFactors = FALSE)
geno  <- read.csv("data/demo-genotypes.csv", row.names = 1)

df <- merge(pheno, covar, by = "sample_id")

common_ids <- intersect(df$sample_id, rownames(geno))
df   <- df[match(common_ids, df$sample_id), ]
geno <- geno[common_ids, , drop = FALSE]

test_snp_lm <- function(x, df){
  if (any(is.na(x))){
    x[is.na(x)] <- mean(x, na.rm = TRUE)
  }
  fit <- lm(trait ~ x + age + sex + PC1 + PC2, data = df)
  co <- summary(fit)$coefficients
  c(
    beta = unname(co["x", "Estimate"]),
    se   = unname(co["x", "Std. Error"]),
    p    = unname(co["x", "Pr(>|t|)"])
  )
}

res <- t(apply(geno, 2, test_snp_lm, df = df))
res <- as.data.frame(res)
res$snp <- rownames(res)
res$logp <- -log10(res$p)
```

---

## Manhattan-Style Plot (Conceptual)

In real GWAS, SNPs are ordered by chromosome and position.

In this demo, we plot SNP index vs –log10(p).

```{r manhattan-plot}
#| message: false
#| warning: false

res$index <- seq_len(nrow(res))
pal <- cdi_palette()

ggplot(res, aes(x = index, y = logp)) +
  geom_point(color = pal$teal, alpha = 0.75, size = 1.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = pal$highlight) +
  labs(
    title = "Manhattan-style overview",
    subtitle = "SNP index vs -log10(p)",
    x = "SNP index",
    y = "-log10(p-value)"
  ) +
  cdi_theme()
```

### Interpretation

Points rising above the dashed line represent nominally significant variants.

However:

- Nominal significance is not genome-wide significance.
- Isolated peaks are more convincing than diffuse elevation.
- Widespread elevation suggests inflation or confounding.

---

## QQ Plot

A QQ plot compares observed p-values to those expected under the null.

```{r qq-plot}
#| message: false
#| warning: false

observed <- sort(res$p)
expected <- ppoints(length(observed))

qq_df <- data.frame(
  expected = -log10(expected),
  observed = -log10(observed)
)

pal <- cdi_palette()

ggplot(qq_df, aes(x = expected, y = observed)) +
  geom_point(color = pal$teal, alpha = 0.7, size = 1.6) +
  geom_abline(intercept = 0, slope = 1, color = pal$highlight, linewidth = 0.9) +
  labs(
    title = "QQ plot of GWAS p-values",
    subtitle = "Observed vs expected under null",
    x = "Expected -log10(p)",
    y = "Observed -log10(p)"
  ) +
  cdi_theme()
```

### Interpretation

If points closely follow the diagonal, the test statistics are well calibrated.

Upward deviation at the tail suggests real signals.

Systematic upward deviation across the range suggests inflation.

---

## Genomic Inflation Factor (Lambda)

We quantify inflation using lambda.

```{r genomic-inflation}
#| message: false
#| warning: false

chisq <- qchisq(1 - res$p, df = 1)
lambda <- median(chisq, na.rm = TRUE) / qchisq(0.5, df = 1)
lambda
```

### Interpretation

- Lambda ≈ 1 indicates well-calibrated statistics.
- Lambda > 1 suggests inflation (structure, confounding, or model misspecification).
- Lambda < 1 suggests overcorrection or conservative testing.

Lambda is a diagnostic, not a verdict.

---

## What This Lesson Adds

Association tables become interpretable when visualized.

We now have tools to detect:

- global inflation
- structured bias
- isolated strong signals
- tail behavior

Only after these diagnostics can we responsibly interpret top hits.

Lesson 6 moves from statistical signal to biological calibration.
